A putative causal map of natural variation across the human phenome

### 2017-08-03 docs/manuscript.rmd
updated

---
title: Learning pleiotropic signatures to construct a causal map of the human phenome
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
header-includes:
   - \usepackage{amsmath}
bibliography: manuscript.bib
csl: vancouver.csl
---

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this, each assuming a different model of horizontal pleiotropy. Here we show that across a range of horizontal pleiotropy models instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. We have integrated these methods into a mixture-of-experts machine learning approach (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome, comprising the evaluation of causality between over a million pairs of traits. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).

@@

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure [@DaveySmithHemani2014]. The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017]. It is considered best practice to report estimates from a range of methods as a sensitivity analysis when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions. However it is also desirable to consolidate across methods under some circumstances. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no HP, it can have high false positive or low true positive rates in the presence of HP compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.

In this paper we introduce two innovations towards improving the reliability of MR estimates. First, we introduce an approach to discard genetic variants that are likely to be invalid. Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection. Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates. Motivated by the recent avalanche [@Hartwig2016] of 2SMR publications, with similar [@Xu2017; @Larsson2017] or contradicting [@Prins2016; @Inoshita2016] conclusions despite using the same data [@Hartwig2016], we developed a graph database to represent these estimates in a consistent manner. We consider this to be a 'working draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward, and that future corrections and refinements will inevitably follow as data grows and 2SMR methods evolve.

### Box 1: Models of pleiotropy

In this paper we separate pleiotropy into two forms for simplicity - vertical and horizontal. Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait. It has also been termed Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998].

By contrast, horizontal pleiotropy can manifest in a myriad of different ways, but the guiding principle is that the genetic variant influences two different phenotypes through independent pathways. Horizontal pleiotropy has also been termed biological pleiotropy [@Solovieff2013], Type I pleiotropy [@Wagner2011], developmental pleiotropy or selectional pleiotropy [@Paaby2013]. From a MR statistical perspective, one can be more specific and state that horizontal pleiotropy occurs where a SNP influences an outcome trait through some pathway other than the exposure trait.

For illustration, Hu et al (2016) (Figure 3) enumerate the biological mechanisms that can lead to pleiotropy: a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016]. Whether we consider each of these mechanisms to be vertical pleiotropy (advantageous for MR) or horizontal pleiotropy (problematic for MR) depends on which phenotypes are being interrogated. For example, consider a scenario where a SNP influences the expression levels of two genes, and the first gene influences trait A while the second gene influences trait B. Then:

- The genetic variant is exhibiting a vertical pleiotropic effect on trait A through the first gene expression level. Here, all other assumptions being met, the genetic variant would be a valid instrument if used to estimate the causal effect of the gene expression level on trait A. However, 
- the SNP is exhibiting a horizontal pleiotropic effect on trait A and trait B through the two gene expression levels. Here the genetic variant would be invalid if used to estimate the causal effect of trait A on trait B.

@@

The simplest approach to selecting instruments for performing MR is to take SNPs that have been declared significant in the published GWAS for the exposure. This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample. These results are often recorded in public GWAS catalogs. Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016]. Complete summary data refers to association results for all SNPs used in a GWAS i.e. not only those passing statistical thresholds for significance. We call this the "tophits" strategy.

@@

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception. Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set. The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy. As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].

Crucially however, we observed in our simulations that none of the available methods were conservative and all, including MR-MoE, had FDR above 0.2. Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion. We illustrated showed how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.

We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates. MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception. Based on our simulations we expect that those associations that we reported to be 'significant' are liable to high type 1 error rates.

One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results, but in general we strongly advocate the use of reporting multiple MR methods to appraise any specific the impact of HP on a causal estimate. In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Such associations can be explained statistically because the instruments proxy the liability for that particular phenotype, not that phenotype's observed manifestation. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS of disease is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic due to the instruments proxying disease liability and not diseae outcomes.

A large proportion of the associations were estimated using only a single instrumental variable, a circumstance in which cause, reverse cause or horizontal pleiotropy cannot be immediately distinguished. Though methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017]. Other problems can also manifest. Frailty effects [@Aalen2015] could induce associations for late-onset traits, and often detailed modelling for how this can influence MR is warranted [@Noyce2017; @Anderson2017]. Genetic variants could strongly relate to several phenotypes, for example SNPs that influence LDL cholesterol are also likely to influence other lipid fractions, making it difficult to ascertain which amongst them is the true causal exposure [@Burgess2014a]. The measured feature for which genetic associations are known could relate in complicated ways to the biological entities that is truly causal [@DaveySmithHemani2014], for example higher circulating levels of the natural antioxidant extracellular superoxide dismutase (EC-SOD) appears to causally relate to higher risk of coronary heart disease [@Juul2003] because the phenotypic measure relates inversely to the EC-SOD levels in arterial walls, hence *in situ* activity is lower.

We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the model selection is data driven. This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance. The extent to which a machine-based method selector incurs higher type 1 error rates more than a human selector is not clear. Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to other MR methods that do not suffer from the issue of data driven model selection.

The field of MR is evolving. The advent of GWAS databases [@Hindorff2010; @Staley2016] and automated 2SMR [@Hemani2016] has trivialised the analytical aspect of investigating specific causal hypotheses. Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal. First, though causality is not guaranteed by MR, it can still be highly informative for confirming or negating specific hypotheses. Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses. The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.

@@

This work was supported by the UK Medical Research Council (MC_UU_12013/1).

>>>>

---
title: A putative causal map of natural variation across the human phenome
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
header-includes:
   - \usepackage{amsmath}
bibliography: manuscript.bib
csl: vancouver.csl
---

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a range of pleiotropy models, instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. Methods have been developed in an attempt to protect MR from different patterns of horizontal pleiotropy, and here we have designed a mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts the most appropriate model to use for any specific causal analysis, improving on both power and false discovery rates. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, comprising the evaluation of causality between over a million pairs of traits. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)), and we use it here to highlight the numerous challenges that remain in automated causal inference.

@@

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy (HP), where they influence the outcome through a pathway other than the exposure (Box 1). The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of HP [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017]. It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017]. However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no HP, it can have high false positive or low true positive rates in the presence of HP compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of HP exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of HP, and therefore an automated method for instrument selection may be necessary.

In this paper we introduce two innovations towards improving the reliability of MR estimates. First, we present an approach to discard genetic variants that are likely to be invalid. Second, we hypothesised that characteristics of the summary data could indicate which method would be most reliable, and we introduce new machine learning approaches that attempt to automate both instrument and method selection. Using curated GWAS summary data for thousands of phenotypes [@Hemani2016], we use these new methods to construct a graph of millions of causal estimates. Motivated by the recent avalanche [@Hartwig2016] of 2SMR publications, with similar [@Xu2017; @Larsson2017] or contradicting [@Prins2016; @Inoshita2016] conclusions despite using the same data [@Hartwig2016], we developed a graph database to represent these estimates in a consistent manner. We consider this to be a 'working draft' of the causal map of the human phenome, but raise caution throughout that its interpretation is far from straightforward, and that future corrections and refinements will inevitably follow as data grows and 2SMR methods evolve.

### Box 1: Pleiotropic mechanisms

Pleiotropy is the phenomenon whereby a single genetic variant influences multiple phenotypes. The mechanisms behind it can arise through many different mechanisms, for example a single variant influences multiple genes; a single gene generates multiple gene products; a single gene product has multiple cellular functions; or a single cellular function influences multiple phenotypes [@Hu2016]. How we classify the pleiotropic mechanism depends on the application, and in this paper we discuss pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy has also been known as mediated pleiotropy [@Solovieff2013], because the genetic variant influences one trait, which in turn influences a second trait. It has also been termed, in some contexts, Type II pleiotropy [@Wagner2011] and secondary pleiotropy [@Hodgkin1998]. It is this form of pleiotropy that MR assumes to make causal inference between two traits.

By contrast, horizontal pleiotropy can manifest in a myriad of different ways, but the basic principle is that the genetic variant influences two different phenotypes through independent pathways. Depending on context, horizontal pleiotropy can also be termed biological pleiotropy [@Solovieff2013], Type I pleiotropy [@Wagner2011], developmental pleiotropy or selectional pleiotropy [@Paaby2013]. From the perspective of MR, horizontal pleiotropy arises when a SNP influences an outcome trait through some pathway other than the exposure trait.

A genetic variant can exhibit both horizontal and vertical pleiotropy in different MR analyses. For example, consider a scenario where a SNP influences the expression levels of two genes (X and Y), and gene X influences trait A while gene Y influences trait B. Then:

- The genetic variant is exhibiting a vertical pleiotropic effect on trait A and the expression level of gene X. Here, all other assumptions being met, the genetic variant would be a valid instrument if used to estimate the causal effect of the gene expression level on trait A. However, 
- the SNP is exhibiting a horizontal pleiotropic effect on trait A and trait B through the two gene expression levels. Here the genetic variant would be invalid if used to estimate the causal effect of trait A on trait B.

![](images/pleiotropy.png)

Adapted from Figure 3 in Hu et al (2016).

@@

The simplest approach to selecting instruments is to take SNPs that have been declared significant in the published GWAS for the exposure. This typically involves obtaining SNPs that surpass $p < 5 \times 10^{-8}$, using clumping to obtain independent SNPs, and then replicating in an independent sample. These results are often recorded in public GWAS catalogs. Alternatively the clumping procedure can be performed using complete summary data in MR-Base [@Hemani2016]. Complete summary data refers to association results for all SNPs used in a GWAS i.e. not only those passing statistical thresholds for significance. We call this the "tophits" strategy.

@@

Motivated by a recent flurry of activity in MR methodology development, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set. The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy. As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto]. We applied the method to the curated set of GWAS summary data present in MR-Base [@Hemani2016], generating several million MR estimates. MR-MoE selected a method that indicated some pattern of horizontal pleiotropy for almost 90% of cases, reinforcing the notion that it is the rule rather than the exception.

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception. Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion. We illustrated how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.

Methods have been introduced recently that seek to improve MR estimates through invalid instrument detection and removal [@Corbin2016; @Verbanck2017; @Zhu2017]. These attack the problem from the perspective that variants which have a relatively large influence on the causal estimate, or provide substantially different estimates from the other variants, are likely to be exhibiting HP, and their removal from the analysis may be warranted. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], because CRP itself is downstream of many cellular functions and most GWAS signals will not be specific for the trait. Steiger filtering has attempted to sidestep this issue in removing invalid instruments, seeking to detect those instruments which have associations with the outcome that are too large to be consistent with the exposure causing the outcome. We acknowledge that unmeasured confounding and measurement error can introduce problems with Steiger filtering [@Hemani2017], but note that our simulations indicate that all individual MR methods improve in performance following its application.

In addition to HP, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Such associations can be explained statistically because the instruments proxy the liability for that particular phenotype, not that phenotype's observed manifestation. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS of disease is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic due to the instruments proxying disease liability and not diseae outcomes.

A large proportion of the associations were estimated using only a single instrumental variable, a circumstance in which cause, reverse cause or horizontal pleiotropy cannot be immediately distinguished. Though methods are emerging to attempt to delineate between models of reverse cause, pleiotropy and multiple causal variants in the same region [@Zhu2016; @Richardson2017], separating vertical from horizontal pleiotropy with a single instrument is not yet possible in the two-sample MR framework though the use of mediation-based approaches may be informative in the one-sample setting [@Wang2017; @Hemani2017].

Other problems, in addition to HP, can also manifest. Frailty effects [@Aalen2015] could induce associations for late-onset traits, and often detailed modelling for how this can influence MR is warranted [@Noyce2017; @Anderson2017]. Genetic variants could strongly relate to several phenotypes, for example SNPs that influence LDL cholesterol are also likely to influence other lipid fractions, making it difficult to ascertain which amongst them is the true causal exposure [@Burgess2014a]. We see this issue manifesting in Figure 4, where there appear to be causal relationships between different lipid fractions. The measured feature for which genetic associations are known could relate in complicated ways to the biological entities that is truly causal [@DaveySmithHemani2014], for example higher circulating levels of the natural antioxidant extracellular superoxide dismutase (EC-SOD) appears to causally relate to higher risk of coronary heart disease [@Juul2003] because the phenotypic measure relates inversely to the EC-SOD levels in arterial walls, hence *in situ* activity is lower [@DaveySmithHemani2014]. Dynastic effects can also confound MR estimates, arising when the exposure trait in a previous generation influences the environment of the current generation. Here, the SNPs being used as instruments for the exposure will be associated with a potential confounder in the analysis [@VonHinke2016]. The associations depicted in Figure 4 involving educational attainment may be susceptible to bias arising from this mechanism, and protecting against it will likely require alternative study designs, for example by using sibling pairs.

We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the model selection is data driven. This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance. One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results. The extent to which a machine-based method selector incurs higher type 1 error rates more than a human selector is not clear. Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to other MR methods that do not suffer from the issue of data driven model selection.

In light of these issues, we emphasise that we do not consider the reporting of MR-MoE (or any other single MR method) to be seen as performing a Mendelian randomization study. Rather, it can be used to motivate further detailed follow up which should include sensitivity analyses [@Hemani2016], incorporating biological knowledge regarding instruments, and triangulation with other sources of evidence or experimental designs [@Lawlor2017]. Demanding a rigorous follow up of putative associations is also essential for avoiding issues that may arise due to cherry picking of MR results. This is especially important when a large repository of putative associations, such as MR-EvE, is made publicly available.

The field of MR is evolving. The advent of GWAS databases [@Hindorff2010; @Staley2016] and automated 2SMR [@Hemani2016] has trivialised the analytical aspect of investigating specific causal hypotheses. Despite the limitations described above, the construction of a causal graph of 'everything versus everything' does have appeal. First, though causality is not guaranteed by MR, it can still be highly informative for supporting or negating specific hypotheses. Indeed, a well-powered MR that negates a hypothesis is likely less fallible to the pitfalls described above than using MR to confirm a positive association [@Vanderweele2014]. Second, it paves the way for new approaches to search for novel putative associations and improve reliability in single analyses. The potential to exploit GWAS summary data within the properties of graph databases to aid with both of these endeavours could be transformative in biological research.

@@

This work was supported by the UK Medical Research Council (MC_UU_12013/1, MC_UU_12013/8, MR/N501906/1); Cancer Research UK (C52724/A20138, C18281/A19169); and the SPHERE Interdisciplinary Research Collaboration funded by the UK Engineering and Physical Sciences Research Council under grant EP/K031910/1
